# Reproducible Research: Peer Assessment 1

---
title: "Reproducible Research - Week 1 course project"
author: "Avinash Anand"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
```

## Loading and preprocessing the data
##### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=TRUE, message=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

if("downloader" %in% rownames(installed.packages()) == FALSE){
  install.packages("downloader")  #installing the package
  library(downloader)
} else{
  library(downloader)
}

#load ggplot2 package
library(ggplot2)

if(file.exists("Proj_dataFile_Week1.zip") == FALSE){
  #downloading the source file using downloader package
  download(url, dest = "Proj_dataFile_Week1.zip", mode = "wb")
}

if(!file.exists("activity.csv")){
  unzip("Proj_dataFile_Week1.zip") #unzip the downloaded file  
}

#Reading the file
actdf <- read.csv("activity.csv", stringsAsFactors = FALSE)
```
##### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r, Processing, echo=TRUE}
actdf <- read.csv("activity.csv", stringsAsFactors = FALSE)
```


## What is mean total number of steps taken per day?
```{r}
agg <- aggregate(steps~date, actdf, FUN = sum, na.action = na.omit)
```

##### 1. Make a histogram of the total number of steps taken each day
```{r}
g<- ggplot(data=agg, aes(steps))
plt <- g + geom_histogram(fill = "firebrick", binwidth = 600) +
  xlab("Total number of steps taken per day") +
  ylab("Frequency")
print(plt)
```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
StepsMean <- mean(agg$steps, na.rm = TRUE)
StepsMeadian <- median(agg$steps, na.rm = TRUE)
```
* Mean: `r StepsMean`
* Median:  `r StepsMeadian`

-----

## What is the average daily activity pattern?
```{r}
avg <- aggregate(steps~interval, actdf, FUN = mean, na.action = na.omit)
```

##### 1. Make a time series plot
```{r}
g <- ggplot(avg, aes(x=interval, y=steps))
g + geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken") 
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
steps <- avg[which.max(avg$steps),]
```

* Most Steps at: `r steps`

----

## Imputing missing values
##### 1. Calculate and report the total number of missing values in the dataset 
```{r}
numMissingValues <- sum(is.na(actdf))
```

* Number of missing values: `r numMissingValues`

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
* Using the mean of the interval for filling the missing values, new dataset is created

##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
new_actdf<- actdf
missVal<- is.na(new_actdf$steps)
avgsteps<-tapply(new_actdf$steps, new_actdf$interval, FUN = mean, na.rm=TRUE, simplify = TRUE)
new_actdf$steps[missVal]<-avgsteps[as.character(new_actdf$interval[missVal])]
```


##### 4. Make a histogram of the total number of steps taken each day 
```{r}
agg1 <- aggregate(steps~date, new_actdf, FUN = sum, na.action = na.omit)
g<- ggplot(data=agg, aes(steps))
g + geom_histogram(fill = "firebrick", binwidth = 1000) + labs(title = "Histogram of Steps per day, including missing values", x = "Steps per day", y = "Frequency")
```

##### ... and Calculate and report the mean and median total number of steps taken per day. 
```{r}
stepsByDayMeanImputed <- mean(agg$steps, na.rm = TRUE)
stepsByDayMedianImputed <- median(agg$steps, na.rm = TRUE)
```
* Mean (Imputed): `r stepsByDayMeanImputed`
* Median (Imputed):  `r stepsByDayMedianImputed`


----

## Are there differences in activity patterns between weekdays and weekends?
##### 1. Create a new factor variable in the dataset with two levels weekday and weekend indicating whether a given date is a weekday or weekend day.

```{r}
days<-c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
actdf$day<-ifelse(is.element(weekdays(as.Date(actdf$date)), days), "Weekday", "Weekend")
```

##### 2. Make a panel plot containing a time series plot

```{r}
avg1 <- aggregate(steps~interval+day, actdf, FUN = mean, na.action = na.omit)
g <- ggplot(data = avg1, aes(y=steps, x=interval, color = day))
g + facet_wrap(~avg1$day, nrow=2, ncol=1) + geom_line() +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```

